---
title: "Waterlogging microbiome analysis"
author: "Hamish McLean"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cerulean
    highlight: tango
params:
  pipeline: "DADA2"
---

```{r setup, include=FALSE}

if (!exists("params")) params <- list(pipeline = "DADA2")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)

# Kable options
options(
  browser = "false",
  kableExtra.html.bsTable = TRUE,
  kableExtra.bootstrap_options = c("striped", "hover", "condensed"),
  kableExtra_view_html = interactive(),
  knitr.table.format = "html",
  viewer = NULL # Disable viewer to avoid opening in browser
)

Sys.setenv("BROWSER" = "false")

```

```{r libraries}

library(buildmer)
library(car)
library(data.table)
library(DESeq2)
library(dplyr)
library(emmeans)
library(ggpubr)
library(gridExtra)
library(here)
library(kableExtra)
library(lme4)
library(lmPerm)
library(partR2)
library(patchwork)
library(performance)
library(permutes)
library(see)
library(vegan)
library(viridis)

```

```{r functions}

here <- here::here()

FUNC_DIR <- here("functions")

source(file.path(FUNC_DIR, "loadme.R"))
source(file.path(FUNC_DIR, "metabarcoding.R"))
source(file.path(FUNC_DIR, "metafuncs.R"))
source(file.path(FUNC_DIR, "rarefaction.R"))

```

```{r constants}

DATA_DIR    <- normalizePath(here("..", "Data"))
FIGURES_DIR <- here("figures")

# Bioinformatics pipeline (defaults to DADA2)
# This should be set in the YAML header or passed as a parameter
PIPELINE    <- params$pipeline # DADA2 or USEARCH
ASV_FILTER  <- 0.001 # ASV filter for removing low abundance ASVs
READ_FILTER <- 0.05  # Read count filter for rarefaction
TAX_CONF    <- 0.65  # Confidence threshold for taxonomic rank assignment
SEED        <- 25647 # Random seed

# Colour blind palette
cbPalette <- c(
  "#000000", "#E69F00", "#56B4E9", "#009E73", 
  "#F0E442", "#0072B2", "#D55E00", "#CC79A7"
)

# Model designs
# Timepoint model for E1
# formula_t <- y ~ treatment * timepoint + (1 | plot) + (1 | block)
T_DESIGN <- ~ treatment * timepoint + (1 | plot)

# Experiment model without timepoint 2
E_DESIGN <- ~ treatment * experiment + (1 | experiment:block)

```

## Load data

```{r load data}

# File extensions based on pipeline
if(PIPELINE == "DADA2") {
  asv_ext <- ".asv_table_filtered.txt"
  fun_tax <- "FUN.sintax.taxa"
  bac_tax <- "BAC.sintax.taxa"
} else if(PIPELINE == "USEARCH") {
  asv_ext <- ".zotu_table.txt"
  fun_tax <- "zFUN.sintax.taxa"
  bac_tax <- "zBAC.sintax.taxa"
} else {
  stop("Pipeline not recognised")
}

# Load fungal data

FUN <- loadData(
  here(DATA_DIR, PIPELINE, paste0("FUN", asv_ext)),
  here(DATA_DIR, "metadata.txt"),
  here(DATA_DIR, PIPELINE, fun_tax),
  tax_conf = TAX_CONF,
  RHB = "FUN"
)

# Load bacterial data

BAC <- loadData(
  here(DATA_DIR, PIPELINE, paste0("BAC", asv_ext)),
  here(DATA_DIR, "metadata.txt"),
  here(DATA_DIR, PIPELINE, bac_tax),
  tax_conf = TAX_CONF,
  RHB = "BAC"
)

# Remove ".raw" from the end of the colnames of the countData
colnames(FUN$countData) <- gsub("\\.raw$", "", colnames(FUN$countData))
colnames(BAC$countData) <- gsub("\\.raw$", "", colnames(BAC$countData))

# Add a new column to colData that combines experiment and timepoint
FUN$colData$exp_tp <- paste(FUN$colData$experiment, FUN$colData$timepoint, sep = "|")
BAC$colData$exp_tp <- paste(BAC$colData$experiment, BAC$colData$timepoint, sep = "|")

# Convert all columns to factors
FUN$colData[] <- lapply(FUN$colData, as.factor)
BAC$colData[] <- lapply(BAC$colData, as.factor)

```

## Filter data

### Filter taxa

Plantae taxa are filtered from fungal `taxData`.
Chloroplast and Eukaryote  taxa are filtered from bacterial `taxData`.
Corresponding ASVs are removed from `countData`.

```{r filter taxa}

# Fungi: Plantae ASVs
cat(
  "Fungi:\n",
  " - Total ASVs:", nrow(FUN$taxData), "\n",
  " - Plantae ASVs:", length(grep("Plantae", FUN$taxData$kingdom)), "\n\n"
)

# Filter Plantae ASVs
filt <- rownames(FUN$taxData[grep("Plantae", FUN$taxData$kingdom), ])

if(length(filt) == 0) {
  cat("No fungal ASVs to remove\n")
} else {
  cat("Fungi: removing", length(filt), "ASVs\n")
  FUN$taxData <- FUN$taxData[!rownames(FUN$taxData) %in% filt, ]
  FUN$countData <- FUN$countData[!rownames(FUN$countData) %in% filt, ]
}

# Bacteria: Chloroplast (Streptophyta) and Eukaryote ASVs
cat(
  "Bacteria:\n",
  " - Total ASVs:", nrow(BAC$taxData), "\n",
  " - Chloroplast ASVs:", nrow(BAC$taxData[
    grepl("Streptophyta", BAC$taxData$genus) & 
    as.numeric(BAC$taxData$g_conf) >= TAX_CONF,
  ]), "\n",
  " - Eukaryote ASVs:", length(grep("Eukaryota", BAC$taxData$kingdom)), "\n\n"
)

# Filter Chloroplast and Eukaryote ASVs
filt <- rownames(
  BAC$taxData[
    grepl("Streptophyta", BAC$taxData$genus) & 
    as.numeric(BAC$taxData$g_conf) >= TAX_CONF,
  ]
)

filt <- unique(c(filt, rownames(BAC$taxData[grep("Eukaryota", BAC$taxData$kingdom), ])))

if(length(filt) == 0) {
  cat("No Bacterial ASVs to remove\n")
} else {
  cat("Bacteria: removing", length(filt), "ASVs\n")
  BAC$taxData   <- BAC$taxData[!rownames(BAC$taxData) %in% filt, ]
  BAC$countData <- BAC$countData[!rownames(BAC$countData) %in% filt, ]
}

```

### Rarefaction

Remove samples with read count below `r READ_FILTER * 100` % of median.

```{r rarefaction}

rare_fun <- gfunc(FUN$countData, FUN$colData, "Fungi")

rare_bac <- gfunc(BAC$countData, BAC$colData, "Bacteria")

rarefaction_plots <- grid.arrange(
  rare_bac, rare_fun,
  # left = textGrob(label = expression("log"[10] * " aligned sequences"), rot = 90),
  bottom = "ASV count", nrow = 2
)

ggsave(filename = "rarefaction_plots.png", plot = rarefaction_plots, path = FIGURES_DIR)

rarefaction_plots

# Fungi
med <- median(colSums(FUN$countData))
filt <- !colSums(FUN$countData) > med * READ_FILTER
cat("Fungi: ", sum(filt), "sample(s) with <", READ_FILTER * 100, "% of median reads\n\n")

# Bacteria
med <- median(colSums(BAC$countData))
filt <- !colSums(BAC$countData) > med * READ_FILTER
cat("Bacteria: ", sum(filt), "sample(s) with <", READ_FILTER * 100, "% of median reads\n\n")

```

### Filter ASVs by read count

Remove ASVs with less than `r ASV_FILTER * 100` % of total reads across all samples.

```{r asv propotions}

asv_propotions <- function(countData, proportion){
  i <- sum(countData)
  y <- rowSums(countData)
  y <- y[order(y, decreasing = T)]
  asvs <- length(y[(cumsum(y) / i <= proportion)])
  return(asvs)
}

proportions <- c(0.5, 0.9, 0.99, 0.999, 1)

top_asvs <- data.table(
  "proportion" = proportions,
  "Fungi" = lapply(proportions, function(x) asv_propotions(FUN$countData, x)),
  "Bacteria" = lapply(proportions, function(x) asv_propotions(BAC$countData, x))
)

top_asvs |> kbl()

# top_asvs %>%
#   kbl() %>%
#   kable_styling("striped", full_width = F)

```

Keep only ASVs with at least 10 reads in at least 2 samples.

```{r filter asvs}

filter_asvs <- function(countData, min_count = 10, min_samples = 2) {
  keep <- rowSums(countData >= min_count) >= min_samples
  names(keep[keep])
}

# Fungi
# keep <- filter_otus(FUN$countData, ASV_FILTER)
keep <- filter_asvs(FUN$countData)
cat(
  "Fungi:\n", 
  " - total ASVs:", nrow(FUN$countData), "\n",
  " - removing", nrow(FUN$countData) - length(keep), "ASVs\n",
  " - remaining ASVs:", length(keep), "\n\n"
)

FUN$taxData <- FUN$taxData[rownames(FUN$taxData) %in% keep,]
FUN$countData <- FUN$countData[rownames(FUN$countData) %in% keep,]

# Bacteria
# keep <-  filter_asvs(BAC$countData, ASV_FILTER)
keep <- filter_asvs(BAC$countData)
cat(
  "Bacteria:\n",
  " - total ASVs:", nrow(BAC$countData), "\n",
  " - removing", nrow(BAC$countData) - length(keep), "ASVs\n",
  " - remaining ASVs:", length(keep), "\n\n"
)

BAC$taxData <- BAC$taxData[rownames(BAC$taxData) %in% keep,]
BAC$countData <- BAC$countData[rownames(BAC$countData) %in% keep,]

```

## Absolute abundance normalisation

ASV normalisation is performed using qPCR theoretical copy number data.
Copy number is calculated per mg of root sample from the qPCR data.

### Prepare qPCR abundance data

```{r qPCR abundance}

abundance <- read.csv(here(DATA_DIR, "abundance.csv"))

FUN$abundance <- subset(abundance, Target == "ITS")
BAC$abundance <- subset(abundance, Target == "16S")

rownames(FUN$abundance) <- FUN$abundance$Sample
rownames(BAC$abundance) <- BAC$abundance$Sample

# Add `copy_number` to colData from abundance data `TCN_mg`
FUN$colData$copy_number <- FUN$abundance[rownames(FUN$colData), "TCN_mg"]
BAC$colData$copy_number <- BAC$abundance[rownames(BAC$colData), "TCN_mg"]

FUN$colData$log10_copy_number <- log10(FUN$colData$copy_number + 1)
BAC$colData$log10_copy_number <- log10(BAC$colData$copy_number + 1)

```

## Subset data

For both fungi and bacteria:
Subset by each question (experiment and timepoint).
Subset by each source (root and soil).

```{r subset data}

FUN_exp <- list(
  root = list(
    countData = FUN$countData[, FUN$colData$source == "root" & FUN$colData$timepoint == 1],
    colData = FUN$colData[FUN$colData$source == "root" & FUN$colData$timepoint == 1, ]
  ),
  soil = list(
    countData = FUN$countData[, FUN$colData$source == "soil" & FUN$colData$timepoint == 1],
    colData = FUN$colData[FUN$colData$source == "soil" & FUN$colData$timepoint == 1, ]
  )
)

FUN_time <- list(
  root = list(
    countData = FUN$countData[, FUN$colData$source == "root" & FUN$colData$experiment == 1],
    colData = FUN$colData[FUN$colData$source == "root" & FUN$colData$experiment == 1, ]
  ),
  soil = list(
    countData = FUN$countData[, FUN$colData$source == "soil" & FUN$colData$experiment == 1],
    colData = FUN$colData[FUN$colData$source == "soil" & FUN$colData$experiment == 1, ]
  )
)

BAC_exp <- list(
  root = list(
    countData = BAC$countData[, BAC$colData$source == "root" & BAC$colData$timepoint == 1],
    colData = BAC$colData[BAC$colData$source == "root" & BAC$colData$timepoint == 1, ]
  ),
  soil = list(
    countData = BAC$countData[, BAC$colData$source == "soil" & BAC$colData$timepoint == 1],
    colData = BAC$colData[BAC$colData$source == "soil" & BAC$colData$timepoint == 1, ]
  )
)

BAC_time <- list(
  root = list(
    countData = BAC$countData[, BAC$colData$source == "root" & BAC$colData$experiment == 1],
    colData = BAC$colData[BAC$colData$source == "root" & BAC$colData$experiment == 1, ]
  ),
  soil = list(
    countData = BAC$countData[, BAC$colData$source == "soil" & BAC$colData$experiment == 1],
    colData = BAC$colData[BAC$colData$source == "soil" & BAC$colData$experiment == 1, ]
  )
)

```

```{r iterator}

#' Generic iterator function
#' @description Applies a function to each subset of data
#' @param data_list A list of data subsets (e.g., FUN or BAC)
#' @param name A string representing the name of the data (e.g, "Fungi" or "Bacteria")
#' @param func A function to apply to each subset
#' @param ... Additional arguments to pass to the function
#' @return A list of results from applying the function to each subset
#'
apply_to_subsets <- function(data_list, func, name = NULL, ...) {
  # Iterate over sources (root, soil)
  result_list <- lapply(names(data_list), function(source_name) {
    data <- data_list[[source_name]]

    # Apply the provided function to the innermost list element and names
    # Pass the subset data, name, source_name, and any extra args
    if (!is.null(name)) {
      result <- func(data, name, source_name, ...)
    } else {
      result <- func(data, ...)
    }

    # Return the result (assuming func returns the modified list element or results)
    return(result)
  })
  names(result_list) <- names(data_list) # Set names for the result list
  return(result_list)
}

```

```{r add taxData}

# Add taxData
add_taxData <- function(data, taxData) {
  data$taxData <- taxData
  data
}

FUN_exp <- apply_to_subsets(FUN_exp, add_taxData, taxData = FUN$taxData)
BAC_exp <- apply_to_subsets(BAC_exp, add_taxData, taxData = BAC$taxData)

FUN_time <- apply_to_subsets(FUN_time, add_taxData, taxData = FUN$taxData)
BAC_time <- apply_to_subsets(BAC_time, add_taxData, taxData = BAC$taxData)

```

## Create DESeq objects

DESeq2 objects are created for each subset of data.

Absolute abundance normalisation using DESeq2 size factors calculated from qPCR-derived copy number.

Size factor values are centred around the mean of the copy number.

```{r DESeq}

create_dds_subset <- function(data, name, source, design = ~ 1) {
  if(identical(colnames(data$countData), rownames(data$colData))) {

    # Create DESeq object
    dds <- DESeqDataSetFromMatrix(
      countData = data$countData,
      colData = data$colData,
      design = design
    )

    # Set size factors based on copy number
    sizeFactors(dds) <- data$colData$copy_number / mean(data$colData$copy_number)
    # Add dds to data list
    data$dds <- dds
    return(data)

    print(paste0(name, ": ", source, " DESeq object created"))
  } else {
    stop(paste0(name, ": ", source, " countData and colData do not match"))
  }
}

FUN_exp <- apply_to_subsets(FUN_exp, create_dds_subset, "Fungi")
FUN_time <- apply_to_subsets(FUN_time, create_dds_subset, "Fungi")

BAC_exp <- apply_to_subsets(BAC_exp, create_dds_subset, "Bacteria")
BAC_time <- apply_to_subsets(BAC_time, create_dds_subset, "Bacteria")

```

## ASV and sample summary

### Read and sample summary

```{r read summary}

#' @title Read summary
#' @description Create a summary of the read counts per sample
#' @param dds DESeq2 object
#' @param norm Logical, whether to normalise the counts
#' @return A data.table with the total, mean, median, max and min reads per sample
#'
read_summary <- function(dds, norm = FALSE) {
  countData <- counts(dds, normalize = norm)
  data.table(
    "Total reads" = sum(countData),
    "Mean reads per sample" = mean(colSums(countData)),
    "Median reads per sample" = median(colSums(countData)),
    "Max reads per sample" = max(colSums(countData)),
    "Min reads per sample" = min(colSums(countData))
  )
}

print_read_summary <- function(data, title) {
  cat("\n", title, "\n")
  data.table(
    "Metric" = c(
      "Total reads", "Mean reads per sample", "Median reads per sample", 
      "Max reads per sample", "Min reads per sample"
    ),
    "Root (Raw)" = unlist(read_summary(data$root$dds, FALSE)),
    "Root (Normalized)" = unlist(read_summary(data$root$dds, TRUE)),
    "Soil (Raw)" = unlist(read_summary(data$soil$dds, FALSE)),
    "Soil (Normalized)" = unlist(read_summary(data$soil$dds, TRUE))
  ) #%>%
    # kbl() %>%
    # kable_styling("striped", full_width = F)
}

print_read_summary(FUN_exp, "Fungi Experiment Read Summary") |> kbl()
print_read_summary(FUN_time, "Fungi Timepoint Read Summary") |> kbl()
print_read_summary(BAC_exp, "Bacteria Experiment Read Summary") |> kbl()
print_read_summary(BAC_time, "Bacteria Timepoint Read Summary") |> kbl()

```

### ASV summary 

```{r asv summary, eval = FALSE}

#' @title ASV summary
#' @description Create a summary of the read counts per ASV
#' @param dds DESeq2 object
#' @param norm Logical, whether to normalise the counts
#' @return A data.table with the total, mean, median, max and min reads per ASV
#'
asv_summary <- function(dds, norm = FALSE) {
  countData <- counts(dds, normalize = norm)
  data.table(
    "Total ASVs" = nrow(countData),
    "Mean reads per ASV" = round(mean(rowSums(countData)), 1),
    "Median reads per ASV" = median(rowSums(countData)),
    "Max reads per ASV" = max(rowSums(countData)),
    "Min reads per ASV" = min(rowSums(countData))
  )
}

print_asv_summary <- function(data, title) {
  cat("\n", title, "\n")
  data.table(
    "Metric" = c(
      "Total ASVs", "Mean reads per ASV", "Median reads per ASV", 
      "Max reads per ASV", "Min reads per ASV"
    ),
    "Root (Raw)" = unlist(asv_summary(data$root$dds, FALSE)),
    "Root (Normalized)" = unlist(asv_summary(data$root$dds, TRUE)),
    "Soil (Raw)" = unlist(asv_summary(data$soil$dds, FALSE)),
    "Soil (Normalized)" = round(unlist(asv_summary(data$soil$dds, TRUE)), 2)
  )
}

print_asv_summary(FUN_exp, "Fungi Experiment ASV Summary") |> kbl()


data.table(
  "Metric" = c(
    "Total ASVs", "Mean reads per ASV", "Median reads per ASV", 
    "Max reads per ASV", "Min reads per ASV"
  ),
  "Fungi (Raw)" = unlist(asv_summary(FUN$dds, FALSE)),
  "Fungi (Normalized)" = unlist(asv_summary(FUN$dds, TRUE)),
  "Bacteria (Raw)" = unlist(asv_summary(BAC$dds, FALSE)),
  "Bacteria (Normalized)" = round(unlist(asv_summary(BAC$dds, TRUE)), 2)
) %>%
  kbl() %>%
  kable_styling("striped", full_width = F)


top_asvs <- function(countData, taxData, n = 10) {
  y <- rowSums(countData)
  y <- y[order(y, decreasing = T)]
  xy <- y / sum(y)
  data.frame(
    counts = y[1:n], 
    proportion = xy[1:n], 
    rank = taxData[names(y)[1:n],]$rank
  )
}

top_asvs(FUN$countData, FUN$taxData, n = 10)

top_asvs(counts(BAC$dds, normalize = TRUE), BAC$taxData, n = 10)

```

## Taxonomy Summary

### Taxonomy identifiable

Proportion of ASVs which can be assigned (with the given confidence) at each taxonomic rank.

```{r taxonomy ASVs}

taxonomy_summary <- function(taxData) {
  tx <- copy(taxData)
  setDT(tx)
  cols <- names(taxData)[9:15]
  tx[, (cols) := lapply(.SD, as.numeric), .SDcols = cols]
  
  data.table(
    rank = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
    "0.8" = round(unlist(lapply(cols, function(col) sum(tx[[col]] >= 0.8) / nrow(tx))), 2),
    "0.65" = round(unlist(lapply(cols, function(col) sum(tx[[col]] >= 0.65) / nrow(tx))), 2),
    "0.5" = round(unlist(lapply(cols, function(col) sum(tx[[col]] >= 0.5) / nrow(tx))), 2)
  )
}

# Fungi

taxonomy_summary(FUN$taxData) %>%
  kbl() %>%
  kable_styling("striped", full_width = F)

# Bacteria

taxonomy_summary(BAC$taxData) %>%
  kbl() %>%
  kable_styling("striped", full_width = F)

```

Proportion of reads which can be assigned to each taxonomic rank.

```{r taxonomy reads, eval = FALSE}

taxonomy_summary_reads <- function(taxData, dds) {
  tx <-taxData[rownames(dds), ]
  nc <- counts(dds, normalize = T)
  ac <- sum(nc)
  cols <- names(tx)[9:15]
  # tx[, (cols) := lapply(.SD, as.factor), .SDcols = cols]

  data.table(
    rank = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
    "0.8" = round(unlist(lapply(cols, function(col) sum(nc[which(as.numeric(tx[[col]]) >= 0.8),]) / ac)), 2),
    "0.65" = round(unlist(lapply(cols, function(col) sum(nc[which(as.numeric(tx[[col]]) >= 0.65),]) / ac)), 2),
    "0.5" = round(unlist(lapply(cols, function(col) sum(nc[which(as.numeric(tx[[col]]) >= 0.5),]) / ac)), 2)
  )
}

# Fungi

# taxonomy_summary_reads(FUN$taxData, FUN$dds) %>%
#   kbl() %>%
#   kable_styling("striped", full_width = F)

# Bacteria

taxonomy_summary_reads(BAC$taxData, BAC$dds) %>%
  kbl() %>%
  kable_styling("striped", full_width = F)

```

### Taxonomy plots

```{r taxonomy plots}

```

## Community size

```{r copy number plots}

copy_number_combined <- rbind(
  FUN$colData %>% mutate(type = "Fungi"),
  BAC$colData %>% mutate(type = "Bacteria")
)

copy_number_timepoint <- ggboxplot(
  filter(FUN$colData, experiment == 1), 
  x = "timepoint", y = "log10_copy_number", 
  fill = "weeks",
  xlab = "Timepoint",
  ylab = "log10 Copy number",
  facet.by = c("source")
)

copy_number_experiment <- ggboxplot(
  filter(FUN$colData, timepoint == 1), 
  x = "experiment", y = "log10_copy_number", 
  fill = "weeks",
  xlab = "Experiment",
  ylab = "log10 Copy number",
  facet.by = c("source")
)

copy_number_plots <- ggarrange(
  copy_number_timepoint,
  copy_number_experiment,
  ncol = 1, nrow = 2,
  common.legend = TRUE, legend = "bottom"
)

ggsave(filename = "FUN_copy_number_plots.png", path = FIGURES_DIR)

copy_number_plots

copy_number_timepoint <- ggboxplot(
  filter(BAC$colData, experiment == 1), 
  x = "timepoint", y = "log10_copy_number", 
  fill = "weeks",
  xlab = "Timepoint",
  ylab = "log10 Copy number",
  facet.by = c("source")
)

copy_number_experiment <- ggboxplot(
  filter(BAC$colData, timepoint == 1), 
  x = "experiment", y = "log10_copy_number", 
  fill = "weeks",
  xlab = "Experiment",
  ylab = "log10 Copy number",
  facet.by = c("source")
)

ggarrange(
  copy_number_timepoint,
  copy_number_experiment,
  ncol = 1, nrow = 2,
  common.legend = TRUE, legend = "bottom"
)

ggsave(filename = "BAC_copy_number_plots.png", path = FIGURES_DIR)

```

```{r copy number models}

copy_number_model <- function(data, name, source, design) {
  cat("\n", name, ":", source, "\n\n")
  model <- lmer(update(design, copy_number ~ .), data$colData)
  # summary(model) |> print()
  Anova(model, type = 3)
}

apply_to_subsets(FUN_exp, copy_number_model, "Fungi Experiment", E_DESIGN)
apply_to_subsets(FUN_time, copy_number_model, "Fungi Timepoint", T_DESIGN)

apply_to_subsets(BAC_exp, copy_number_model, "Bacteria Experiment", E_DESIGN)
apply_to_subsets(BAC_time, copy_number_model, "Bacteria Timepoint", T_DESIGN)

```

## Alpha diversity

```{r calculate alpha diversity}

#' @title Calculate alpha diversity
#' @description Calculate alpha diversity metrics for each subset of data
#' @param data A list containing countData and colData
#' @return A list with the alpha diversity data added
#' 
calculate_alpha_diversity <- function(data) {
  alpha_dt <- plot_alpha(data$countData, data$colData, returnData = TRUE)
  data$alphaData <- alpha_dt[as.data.table(data$colData, keep.rownames = "Samples"), on = "Samples"]
  return(data)
}

FUN_exp <- apply_to_subsets(FUN_exp, calculate_alpha_diversity)
FUN_time <- apply_to_subsets(FUN_time, calculate_alpha_diversity)
BAC_exp <- apply_to_subsets(BAC_exp, calculate_alpha_diversity)
BAC_time <- apply_to_subsets(BAC_time, calculate_alpha_diversity)

```

```{r alpha diversity analysis, eval = FALSE}

#' @title Analyze alpha diversity
#' @description Perform alpha diversity analysis on each subset of data
#' @param data A list containing countData and colData
#' @param name A string representing the name of the data (e.g., "Fungi" or "Bacteria")
#' @param source A string representing the source of the data (e.g., "root" or "soil")
#' @param metrics A vector of alpha diversity metrics to analyze
#' @param formula A formula for the model to use in the analysis
#' @return A list with the alpha diversity results
#'
analyse_alpha_diversity <- function(data, name, source, metrics, formula) {
  cat(paste0("\nAnalyzing alpha diversity for ", name, ":", source, "\n"))

  # Run alpha diversity analysis for each metric
  anova_results <- list()
  for (met in metrics) {
    cat("  - Metric:", met, "\n")
    # Ensure the metric column exists and is numeric before ranking
    if (!met %in% names(data$alphaData)) {
      warning("Metric '", met, "' not found in alpha data for ", name, ": ", source)
      next
    }
    if (!is.numeric(data$alphaData[[met]])) {
      warning("Metric '", met, "' is not numeric in alpha data for ", name, ": ", source, ". Skipping ANOVA.")
      next
    }
    
    # Create a temporary column for the ranked metric for the model
    temp_dt <- copy(data$alphaData) # Work on a copy to avoid modifying original alpha_dt in place with 'metric' column
    temp_dt[, model_metric := rank(get(met))]
    # Update the formula to use the temporary 'model_metric'
    model_formula <- update(formula, model_metric ~ .)
    # Ensure all variables in the formula exist in the data
    formula_vars <- all.vars(model_formula)
    if (!all(formula_vars %in% names(temp_dt))) {
      missing_vars <- setdiff(formula_vars, names(temp_dt))
      warning("Missing variables in data for ", name, ":", source, ":", paste(missing_vars, collapse = ", "), ". Skipping ANOVA for metric ", met)
      next
    }
    # Run the permutation model
    res <- tryCatch({
      perm.lmer(model_formula, temp_dt, type = "anova", nperm = 1000)
    }, error = function(e) {
      warning("Error running perm.lmer for ", name, ":", source, ", metric ", met, ": ", e$message)
      NULL # Return NULL on error
    })
    if (!is.null(res)) {
      print(res) # Print the ANOVA table
      # print()
      # anova_results[[met]] <- res$anova # Store the result
    } else {
      # anova_results[[met]] <- "Error" # Indicate error
      warning("ANOVA for ", name, ":", source, ", metric ", met, " failed.")
    }
  }
  # Return results (e.g., the alpha data and ANOVA results)
  # return(data$alphaData) # Return the alpha data with metrics
}

# Define metrics
metrics <- c("S.chao1", "shannon", "simpson")

apply_to_subsets(FUN_exp, analyse_alpha_diversity, "Fungi Experiment", metrics, E_DESIGN)
apply_to_subsets(FUN_time, analyse_alpha_diversity, "Fungi Timepoint", metrics, T_DESIGN)

apply_to_subsets(BAC_exp, analyse_alpha_diversity, "Bacteria Experiment", metrics, E_DESIGN)
apply_to_subsets(BAC_time, analyse_alpha_diversity, "Bacteria Timepoint", metrics, T_DESIGN)

```

### Alpha diversity plots

```{r alpha diversity plots}

measures <- c("Shannon", "Simpson")

# Root fungi

alpha_FUN_root_exp <- plot_alpha(
  FUN_exp$root$countData, FUN_exp$root$colData,
  design = "experiment", colour = "weeks",
  measures = measures,
  type = "box"
) + scale_colour_manual(values = cbPalette) +
  # theme(axis.title.x = element_blank()) +
  ggtitle("Fungi: root")
alpha_FUN_root_exp
ggsave(filename = "alpha_FUN_root_exp.png", path = FIGURES_DIR)

# Soil fungi

alpha_FUN_soil_exp <- plot_alpha(
  FUN_exp$soil$countData, FUN_exp$soil$colData,
  design = "experiment", colour = "weeks",
  measures = measures,
  type = "box"
) + scale_colour_manual(values = cbPalette) +
  # theme(axis.title.x = element_blank()) +
  ggtitle("Fungi: soil")
alpha_FUN_soil_exp
ggsave(filename = "alpha_FUN_soil_exp.png", path = FIGURES_DIR)

# Root bacteria

alpha_BAC_root_exp <- plot_alpha(
  BAC_exp$root$countData, BAC_exp$root$colData,
  design = "experiment", colour = "weeks",
  measures = measures,
  type = "box"
) + scale_colour_manual(values = cbPalette) +
  # theme(axis.title.x = element_blank()) +
  ggtitle("Bacteria: root")
alpha_BAC_root_exp
ggsave(filename = "alpha_BAC_root_exp.png", path = FIGURES_DIR)

# Soil bacteria

alpha_BAC_soil_exp <- plot_alpha(
  BAC_exp$soil$countData, BAC_exp$soil$colData,
  design = "experiment", colour = "weeks",
  measures = measures,
  type = "box"
) + scale_colour_manual(values = cbPalette) +
  # theme(axis.title.x = element_blank()) +
  ggtitle("Bacteria: root")
alpha_BAC_root_exp
ggsave(filename = "alpha_BAC_root_exp.png", path = FIGURES_DIR)

# Soil bacteria

alpha_BAC_soil_exp <- plot_alpha(
  BAC_exp$soil$countData, BAC_exp$soil$colData,
  design = "experiment", colour = "weeks",
  measures = measures,
  type = "box"
) + scale_colour_manual(values = cbPalette) +
  # theme(axis.title.x = element_blank()) +
  ggtitle("Bacteria: soil")
alpha_BAC_soil_exp
ggsave(filename = "alpha_BAC_soil_exp.png", path = FIGURES_DIR)

# alpha_exp_combined <- ggarrange(
#   list(
#     alpha_FUN_root_exp, alpha_FUN_soil_exp,
#     alpha_BAC_root_exp, alpha_BAC_soil_exp
#   ),
#   ncol = 2, nrow = 2,
#   labels = c("Fungi Root", "Fungi Soil", "Bacteria Root", "Bacteria Soil"),
#   common.legend = TRUE, legend = "bottom"
# )

alpha_exp_combined <- (
  alpha_FUN_root_exp + alpha_FUN_soil_exp
) / (
  alpha_BAC_root_exp + alpha_BAC_soil_exp
) +
  plot_layout(guides = "collect") & # Collect legends into a single common legend
  theme(legend.position = "bottom") # Position the common legend


alpha_exp_combined
ggsave(
  filename = "alpha_exp_combined.png", plot = alpha_exp_combined, path = FIGURES_DIR
  # width = 10, height = 10, units = "cm"
)

```

### Permutation ANOVA on alpha diversity index ranks

```{r alpha diversity anova, eval = FALSE}

# Alpha diversity data

fun_alpha <- plot_alpha(FUN$countData, FUN$colData, returnData = TRUE)
bac_alpha <- plot_alpha(BAC$countData, BAC$colData, returnData = TRUE)

# Join with metadata
fun_alpha <- fun_alpha[as.data.table(FUN$colData, keep.rownames = "Samples"), on = "Samples"]
bac_alpha <- bac_alpha[as.data.table(BAC$colData, keep.rownames = "Samples"), on = "Samples"]

combined_alpha <- rbind(
  fun_alpha[, Type := "Fungi"], 
  bac_alpha[, Type := "Bacteria"]
)

# fun_alpha <-  

# Full model
formula <- metric ~ treatment + timepoint + treatment:timepoint + experiment + treatment:experiment + (year + 1 | plot / experiment) + (1 | block / experiment)

# Timepoint model for E1
formula_t <- metric ~ treatment * timepoint + (1 | plot) + (1 | block)

# formula_t <- metric ~ treatment * timepoint + (1 | block / plot)

# Experiment model without timepoint 2
formula_e <- metric ~ treatment * experiment + (1 | block / experiment)


alpha_anova <- function(data, formula, metrics, nperm = 1000) {
  sources <- unique(data$source)
  for (src in sources) {
    data_src <- data[source == src]
    for (met in metrics) {
      data_src[, metric := rank(data_src[[met]])]
      cat("\n", src, ":", met, "\n", sep = "")
      res <- perm.lmer(formula, data_src, type = "anova", nperm = nperm)
      print(res)
    }
  }
}


metrics <- c("S.chao1", "shannon", "simpson")

# Fungi

alpha_anova(
  fun_alpha, formula, metrics, nperm = 1000
)

# Bacteria

alpha_anova(
  bac_alpha, formula, metrics, nperm = 1000
)

```

## Beta diversity

### Bray-Curtis

#### ADONIS

Block included as a main effect in the design formula.

```{r adonis}

e_design <- vg ~ treatment * experiment + experiment:block

t_design <- vg ~ treatment * timepoint

adonis_analysis <- function(data, name, source, design, strata = NULL, nperm = 1000) {
  cat("\n", name, ":", source, "\n")
  set.seed(SEED)
  vg <- vegdist(t(counts(data$dds, normalize = T)), method = "bray")
  adonis2(
    update(design, vg ~ .),
    data$colData,
    strata = if (!is.null(strata)) data$colData[[strata]] else NULL,
    # strata = colData(dds)$experiment:colData(dds)$block, # For repeat measures
    permutations = nperm,
    by = "terms"
  )
}


# Apply adonis analysis to each subset of data
apply_to_subsets(FUN_exp, adonis_analysis, "Fungi Experiment", e_design)
apply_to_subsets(BAC_exp, adonis_analysis, "Bacteria Experiment", e_design)

apply_to_subsets(FUN_time, adonis_analysis, "Fungi Timepoint", t_design, "plot")
apply_to_subsets(BAC_time, adonis_analysis, "Bacteria Timepoint", t_design, "plot")

# df <- result %>% data.frame()
# df$Perc.Var <- df$SumOfSqs / df["Total", "SumOfSqs"] * 100
# df %>%
#   kbl() %>%
#   kable_styling("striped", full_width = F)


```

#### NMDS

```{r nmds}

set.seed(SEED)

#' @title NMDS plot
#' @description Create a NMDS plot for a given DESeq2 object
#' @param dds DESeq2 object
#' @param shape A string representing factor to use for the shape aesthetic
#' @param filename A string representing the filename to save the plot
#' @return A ggplot object of the NMDS plot
#'
nmdsPlot <- function(dds, shape, filename) {
  vg <- vegdist(t(counts(dds, normalize = T)), method = "bray")
  ord <- metaMDS(vg, trace=0)

  plot <- plotOrd(
    scores(ord), 
    colData(dds), 
    design = "treatment", 
    shape = shape, 
    alpha = 0.75, cbPalette = T
  )
  ggsave(filename, path = FIGURES_DIR)
  return(plot)
}

# Experiment

nmds_exp_fun_root <- nmdsPlot(FUN_exp$root$dds, shape = "experiment", filename = "NMDS_FUN_root_exp.png")
nmds_exp_fun_soil <- nmdsPlot(FUN_exp$soil$dds, shape = "experiment", filename = "NMDS_FUN_soil_exp.png")
nmds_exp_bac_root <- nmdsPlot(BAC_exp$root$dds, shape = "experiment", filename = "NMDS_BAC_root_exp.png")
nmds_exp_bac_soil <- nmdsPlot(BAC_exp$soil$dds, shape = "experiment", filename = "NMDS_BAC_soil_exp.png")

nmds_exp_combined <- ggarrange(
  plotlist = list(nmds_exp_fun_root, nmds_exp_fun_soil, nmds_exp_bac_root, nmds_exp_bac_soil),
  common.legend = TRUE,
  legend = "bottom",
  labels = c("Fungi: roots", "Fungi: soil", "Bacteria: roots", "Bacteria: soil")
)
nmds_exp_combined
ggsave("NMDS_exp.png", plot = nmds_exp_combined, path = FIGURES_DIR)

# Timepoint

nmds_time_fun_root <- nmdsPlot(FUN_time$root$dds, shape = "timepoint", filename = "NMDS_FUN_root_time.png")
nmds_time_fun_soil <- nmdsPlot(FUN_time$soil$dds, shape = "timepoint", filename = "NMDS_FUN_soil_time.png")
nmds_time_bac_root <- nmdsPlot(BAC_time$root$dds, shape = "timepoint", filename = "NMDS_BAC_root_time.png")
nmds_time_bac_soil <- nmdsPlot(BAC_time$soil$dds, shape = "timepoint", filename = "NMDS_BAC_soil_time.png")

nmds_time_combined <- ggarrange(
  plotlist = list(nmds_time_fun_root, nmds_time_fun_soil, nmds_time_bac_root, nmds_time_bac_soil),
  common.legend = TRUE,
  legend = "bottom",
  labels = c("Fungi: roots", "Fungi: soil", "Bacteria: roots", "Bacteria: soil")
)
nmds_time_combined
ggsave("NMDS_time.png", plot = nmds_time_combined, path = FIGURES_DIR)


```

### PCA

```{r pca, eval = FALSE}

n_pcs <- 10

# fun_pca <- des_to_pca(FUN$dds)
# # fun_pca <- t(data.frame(t(pca$x) * pca$percentVar))

# bac_pca <- des_to_pca(BAC$dds)
# # bac_pca <- t(data.frame(t(pca$x) * pca$percentVar))

# Apply PCA to subsets
pca_subset <- function(data, name, source) {
  # Calculate PCA scores
  pca <- des_to_pca(data$dds)

  # Print PCA percent variance table
  pca_var <- data.frame(
    PC = paste0("PC", 1:n_pcs),
    perc_var = round(pca$percentVar[1:n_pcs] * 100, 1)
  )
  cat("\n", name, ":", source, "\n\n")
  print(pca_var)

  # Adjust PCA scores with total variance
  data$pca <- pca #t(data.frame(t(pca$x) * pca$percentVar))
  return(data)
}

FUN_exp  <- apply_to_subsets(FUN_exp,  pca_subset, "Fungi experiment")
BAC_exp  <- apply_to_subsets(BAC_exp,  pca_subset, "Bacteria experiment")
FUN_time <- apply_to_subsets(FUN_time, pca_subset, "Fungi timepoint")
BAC_time <- apply_to_subsets(BAC_time, pca_subset, "Bacteria timepoint")



# Summarise PCA
pca_summary_subset <- function(data, design, n_pcs = 10) {
  pcs      <- 1:n_pcs
  pca_data <- as.data.frame(data$pca$x[, pcs])
  pc_names <- colnames(pca_data)
  variance <- data$pca$percentVar[pcs]

  results <- lapply(seq_along(pcs), function(i) {
    pc_name  <- pc_names[i]
    df       <- cbind(PC = pca_data[[pc_name]], data$colData)
    model    <- lm(update(design, PC ~ .), df)
    anova    <- Anova(model, type = 3)
    anova_df <- anova |> data.frame()
    total_ss <- sum(anova_df$Sum.Sq)
    data.table(
      pc      = pc_name,
      factor  = rownames(anova_df),
      SS      = anova_df$Sum.Sq,
      df      = anova_df$Df,
      F       = anova_df$F.value,
      P       = anova_df$Pr..F.,
      var     = anova_df$Sum.Sq / total_ss * 100,
      var_adj = (anova_df$Sum.Sq / total_ss * 100) * variance[i]
    )
  })

  bind_rows(results)[P <= 0.05 & factor != "(Intercept)"]
}


e_design <- ~ treatment * experiment + experiment:block

apply_to_subsets(FUN_exp, pca_summary_subset, name = NULL, e_design)
apply_to_subsets(BAC_exp, pca_summary_subset, name = NULL, e_design)

t_design <- ~ treatment * timepoint

apply_to_subsets(FUN_time, pca_summary_subset, name = NULL, t_design)
apply_to_subsets(BAC_time, pca_summary_subset, name = NULL, t_design)


# Summarise PCA lmer
pca_summary_subset_lmer <- function(data, design, n_pcs = 10) {
  pcs      <- 1:n_pcs
  pca_data <- as.data.frame(data$pca$x[, pcs])
  pc_names <- colnames(pca_data)
  variance <- data$pca$percentVar[pcs]

  results <- lapply(seq_along(pcs), function(i) {
    pc_name  <- pc_names[i]
    df       <- cbind(PC = pca_data[[pc_name]], data$colData)
    message  <- NULL
    model    <- withCallingHandlers(
      lmer(update(design, PC ~ .), df),
      message = function(m) {
        msg <- conditionMessage(m)
        message <<- sub(":.*", "", msg)
        invokeRestart("muffleMessage")
      }
    )
    anova    <- Anova(model, type = 3) |> data.frame()
    # r2       <- r2_nakagawa(model)
    # part_r2 <- partR2(model, partvars = c("treatment", "timepoint", "treatment:timepoint"))
    # print(part_r2)
    data.table(
      pc      = pc_name,
      factor  = rownames(anova),
      Chisq   = anova$Chisq,
      df      = anova$Df,
      P       = anova$Pr..Chisq.,
      # R2_marg = r2$marginal,
      # R2_cond = r2$conditional,
      # R2_part = part_r2$R2,
      message = message
    )
  })

  bind_rows(results)[P <= 0.05 & factor != "(Intercept)"]
}


apply_to_subsets(FUN_exp, pca_summary_subset_lmer, name = NULL, E_DESIGN)
apply_to_subsets(BAC_exp, pca_summary_subset_lmer, name = NULL, E_DESIGN)

apply_to_subsets(FUN_time, pca_summary_subset_lmer, name = NULL, T_DESIGN)
apply_to_subsets(BAC_time, pca_summary_subset_lmer, name = NULL, T_DESIGN)


```

#### PC loadings

PCs of interest:

Experiment
Fungi root: PC8, PC9
Fungi soil: PC6
Bacteria root: PC7
Bacteria soil: PC4, PC8

```{r pc loadings}

# PC loadings
pc_loadings <- function(data, pc, taxData, top_n = 10) {
  df <- cbind(PC = data$pca$x[, pc], data$colData)
  aggregate(PC ~ treatment, df, mean) |> print()

  loadings <- data$pca$rotation[, pc]
  top_asvs <- names(sort(abs(loadings), decreasing = TRUE))[1:top_n]
  data.table(
    PC = pc,
    ASVs = top_asvs,
    loading = loadings[top_asvs],
    taxa = taxData[top_asvs, ]$rank
  )
}

# Fungi root PC8
pc_loadings(FUN_exp$root, 8, FUN$taxData, 10)

# Fungi root PC8
pc_loadings(FUN_exp$root, 8, FUN$taxData, 10)

# Fungi soil PC6
pc_loadings(FUN_exp$soil, 6, FUN$taxData, 10)


# Bacteria root PC7
pc_loadings(BAC_exp$root, 7, BAC$taxData, 10)

# Bacteria soil PC4
pc_loadings(BAC_exp$soil, 4, BAC$taxData, 10)

# Bacteria soil PC8
pc_loadings(BAC_exp$soil, 8, BAC$taxData, 10)

```

Note: Multiply absolute loading by proportion of variance in PC explained by treatment
can then multiply this by proportion of variance of PC to the total community


### Differential abundance

```{r differential abundance}

# Run DESeq model
deseq_model <- function(data, design) {
  design(data$dds) <- design
  data$dds <- DESeq(data$dds)
  data
}

e_design <- ~ treatment * experiment + experiment:block

FUN_exp <- apply_to_subsets(FUN_exp, deseq_model, name = NULL, e_design)
BAC_exp <- apply_to_subsets(BAC_exp, deseq_model, name = NULL, e_design)


# DESeq differential abundance analysis
diff_abundance <- function(data) {
  # Extract results for E1 (reference), E2, and interaction
  res_e1  <- results(data$dds, name = "treatment_1_vs_0")
  res_e2  <- results(data$dds, contrast = list(c("treatment_1_vs_0", "treatment1.experiment2")))
  res_int <- results(data$dds, name = "treatment1.experiment2")

  # Helper function for significance stars
  p_stars <- function(p) {
    ifelse(is.na(p), "",
      ifelse(p < 0.001, "***",
        ifelse(p < 0.01, "**",
          ifelse(p < 0.05, "*", "")
        )
      )
    )
  }

  # Combine into a data.table
  dt <- data.table(
    ASV = rownames(res_e1),
    log2FC_E1 = res_e1$log2FoldChange,
    padj_E1 = res_e1$padj,
    sig_E1 = p_stars(res_e1$padj),
    log2FC_E2 = res_e2$log2FoldChange,
    padj_E2 = res_e2$padj,
    sig_E2 = p_stars(res_e2$padj),
    log2FC_inter = res_int$log2FoldChange,
    padj_inter = res_int$padj,
    sig_inter = p_stars(res_int$padj),
    taxa = data$taxData[rownames(res_e1), ]$rank
  )
  
  # Filter for any significant result
  dt_filt <- dt[
    (padj_E1 < 0.05 | padj_E2 < 0.05 | padj_inter < 0.05) & 
    !is.na(padj_E1) & !is.na(padj_E2) & !is.na(padj_inter)
  ]

  # Optionally, sort by the smallest padj
  dt_filt <- dt_filt[order(pmin(log2FC_E1, log2FC_E2, na.rm = TRUE), decreasing = TRUE)]
  
  return(dt_filt)
}


apply_to_subsets(FUN_exp, diff_abundance)
apply_to_subsets(BAC_exp, diff_abundance)


```