---
title: "Waterlogging qPCR Analysis"
output: 
  html_document:
    mathjax: default
    df_print: kable
    highlight: haddock
---

```{r setup, include = FALSE}

```

```{r libraries, include = FALSE}

library(data.table)
library(ggpubr)
library(here)
library(kableExtra)
library(knitr)
library(latex2exp)
library(rmarkdown)

```

Load qPCR data from csv files exported from the qPCR machine.
Separate the qPCR data into samples, standards, and NTCs.
Remove outliers from the triplicates based on a standard deviation threshold and maximum $C_q$ value.
Caclulate standard curves for the standards on each plate.
Calculate efficiency of each qPCR reactions.

```{r variables}

cq_threshold <- 0.5 # Threshold for difference between triplicate Cq values
max_cq       <- 35 # Maximum Cq value to keep

here         <- here::here() # Set working directory to the location of this file
data_dir     <- file.path(here, "..", "Data") # Data directory

```

## Load qPCR Data

Load qPCR data from csv files exported from the qPCR machine.
Store data from each plate in a list containing data.tables for samples, standards, and NTCs.

```{r data}
# Define the columns to select from the csv file
data_columns <- c(
  "Well", "Target", "Content", "Replicate", "Sample", "Dilution",
  "Cq", "Starting Quantity (SQ)"
)

# Define the plate files
plate_files <- list(
  plate_1 = "Data/2025-02-26 plate 1 roots ITS.csv",
  plate_2 = "Data/2025-03-06 plate 2 roots 16S.csv",
  plate_3 = "Data/2025-03-10 plate 3 soil ITS.csv",
  plate_4 = "Data/2025-03-17 plate 4 soil 16S.csv"
)

# Define sample sources
roots <- list(01:10, 21:30, 41:50)
soil  <- list(11:20, 31:40, 51:60)

#' Load qPCR data from a csv file
#' @param file_path The path to the csv file
#' @param header_rows The number of header rows to skip
#' @param columns The columns to select
#' @return A list containing the samples, standards, and NTCs as data.tables
#'
load_qpcr_data <- function(
  data_dir, filename, plate,
  header_rows = 19, columns = data_columns
) {
  file_path <- file.path(data_dir, filename)

  # Read the csv file and skip the header rows
  data <- fread(file_path, skip = header_rows, select = columns)
  setnames(data, "Starting Quantity (SQ)", "Concentration")

  # Add plate number to the data
  data[, Plate := plate]

  # Rename Content values
  data[, Content := fcase(
    startsWith(Content, "Unkn"), "Sample",
    startsWith(Content, "Std"), "Standard",
    startsWith(Content, "NTC"), "NTC"
  )]

  # Add Type column based on Content
  data[Content == "Sample", Type := fcase(
    Sample %in% unlist(roots), "Root",
    Sample %in% unlist(soil), "Soil"
  )]
  data[Content == "Standard", Type := fcase(
    startsWith(Sample, "Sr"), "Root",
    startsWith(Sample, "Ss"), "Soil"
  )]

  # Add leading 0 to samples for sorting
  data[Content == "Sample", Sample := sprintf("%02d", as.numeric(Sample))]

  # Calculate sample concentration
  data[, Dilution := as.numeric(Dilution)]
  data[Content == "Sample", Concentration := 1 / as.numeric(Dilution)]
  data[Content == "Standard", Dilution := 1 / as.numeric(Concentration)]
  data[, log10_concentration := log10(Concentration)]

  # Construct the plate object
  list(
    plate     = plate,
    data 	 	  = data,
    samples   = data[Content == "Sample"],
    standards = data[Content == "Standard"],
    ntcs      = data[Content == "NTC"]
  )

}

plate_1 <- load_qpcr_data(data_dir, "2025-02-26 plate 1 roots ITS.csv", 1)
plate_2 <- load_qpcr_data(data_dir, "2025-03-06 plate 2 roots 16S.csv", 2)
plate_3 <- load_qpcr_data(data_dir, "2025-03-10 plate 3 soil ITS.csv",  3)
plate_4 <- load_qpcr_data(data_dir, "2025-03-17 plate 4 soil 16S.csv",  4)

# plates <- lapply(
#   plate_files,
#   function(x) load_qpcr_data(x, as.numeric(substr(names(x), nchar(x), nchar(x))))
# )

kable(head(plate_4$samples))

```

## Remove triplicate outliers

Remove outliers with $\Delta C_q >$ `r cq_threshold` between each triplicate and the median.
Samples above a maximum $C_q$ value are also removed.
If only a single replicate from a triplicate remains, the sample is removed and 
the corresponding sample from the other dilution is also removed because 
efficiency cannot be calculated.

Removed outliers are listed in a dataframe.

```{r flag_outliers}

#' Flag outliers from a qPCR plate with Cq standard deviation > threshold
#' or the Cq value above a maximum value or na Cq values.
#' If only a single replicate from a triplicate remains, the sample is flagged.
#' Corresponding samples from the other dilution are also flagged.
#' @param data A data.table containing qPCR data
#' @param threshold The standard deviation threshold
#' @param max_cq The maximum Cq value to keep
#' @return The data.table with a column "Outlier" containing the flag
#'
flag_outliers <- function(data, threshold = cq_threshold, cq_max = max_cq) {
  outliers <- copy(data)

  # Group by replicate and flag outliers
  outliers[,
    Outlier := fcase(
      abs(Cq - median(Cq)) > threshold, "outlier",
      Cq > cq_max, "high Cq",
      is.na(Cq), "missing Cq",
      default = "ok"
    ),
    by = .(Target, Content, Replicate)
  ]

  # Remove singletons
  outliers[,
    Outlier := ifelse(
      Outlier == "ok" & sum(Outlier == "ok") == 1,
      "singleton", Outlier
    ),
    by = .(Target, Content, Replicate)
  ]

  # Remove corresponding samples from other dilutions
  outliers[,
    Outlier := ifelse(
      Outlier == "ok" & sum(Outlier == "ok") <= .N / 2,
      "insufficient data", Outlier
    ),
    by = .(Target, Content, Sample)
  ]

  return(outliers)
}

# Flaq outliers
plate_1$data <- flag_outliers(plate_1$data)
kable(plate_1$data[Outlier != "ok"][order(Sample)], digits = 2)

plate_2$data <- flag_outliers(plate_2$data)
kable(plate_2$data[Outlier != "ok"][order(Sample)], digits = 2)

plate_3$data <- flag_outliers(plate_3$data)
kable(plate_3$data[Outlier != "ok"][order(Sample)], digits = 2)

plate_4$data <- flag_outliers(plate_4$data)
kable(plate_4$data[Outlier != "ok"][order(Sample)], digits = 2)

# Remove outliers
plate_1$samples   <- plate_1$data[Content == "Sample"   & Outlier == "ok"]
plate_1$standards <- plate_1$data[Content == "Standard" & Outlier == "ok"]

plate_2$samples   <- plate_2$data[Content == "Sample"   & Outlier == "ok"]
plate_2$standards <- plate_2$data[Content == "Standard" & Outlier == "ok"]

plate_3$samples   <- plate_3$data[Content == "Sample"   & Outlier == "ok"]
plate_3$standards <- plate_3$data[Content == "Standard" & Outlier == "ok"]

plate_4$samples   <- plate_4$data[Content == "Sample"   & Outlier == "ok"]
plate_4$standards <- plate_4$data[Content == "Standard" & Outlier == "ok"]

```

## Efficiency

Fit a linear model to $C_q \sim \log_{10}(concentration)$ for each sample in a qPCR plate.
Calculate PCR efficiency based on the equation 
$$E = 10^{-1/slope}$$
where $slope$ is the slope of the linear model.
Perfect efficiency is 2.

```{r efficiency}

#' Calculate the efficiency of a qPCR reaction based on the equation
#' \deqn{E = 10^{-1/slope}}
#' @param slope The slope between the Cq and log10 concentration
#' @return The efficiency of the qPCR reaction
#'
efficiency <- function(slope) {
  10^(-1 / slope)
}

#' Calculate summary statistics for each sample in a qPCR plate
#' @param data A data.table containing qPCR data
#' @return A data.table containing the summary statistics
#'
calculate_summary_statistics <- function(data) {
  data[,
    {
      model <- lm(Cq ~ log10(Concentration))
      intercept <- model$coefficients[1]
      slope <- model$coefficients[2]
      list(
        Target     = Target[1],
        N          = .N,
        Intercept  = intercept,
        Slope      = slope,
        R2         = summary(model)$r.squared,
        Efficiency = efficiency(slope)
      )
    },
    by = Sample
  ][order(as.numeric(Sample))]
}

# Calculate summary statistics
plate_1$stats <- calculate_summary_statistics(plate_1$samples)
plate_2$stats <- calculate_summary_statistics(plate_2$samples)
plate_3$stats <- calculate_summary_statistics(plate_3$samples)
plate_4$stats <- calculate_summary_statistics(plate_4$samples)

kable(plate_1$stats, digits = 2)
kable(plate_2$stats, digits = 2)
kable(plate_3$stats, digits = 2)
kable(plate_4$stats, digits = 2)

# Add stats to samples
plate_1$samples <- merge(plate_1$samples, plate_1$stats, c("Sample", "Target"))
plate_2$samples <- merge(plate_2$samples, plate_2$stats, c("Sample", "Target"))
plate_3$samples <- merge(plate_3$samples, plate_3$stats, c("Sample", "Target"))
plate_4$samples <- merge(plate_4$samples, plate_4$stats, c("Sample", "Target"))

```

### Efficiency adjusted $C_q$

$C_q$ can be adjusted for efficiency by the equation:

$$
C_{q, adjusted}=\log_{2}\!\left(E^{C_{q}}\right)
$$

Where $E$ is the PCR efficiency of the sample.

To calculate adjusted copy number, a new standard curve is fit to the adjusted $C_q$ values of the standards.

```{r efficiency_adjustment}

#' Calculate the efficiency adjusted Cq values using the equation
#' \deqn{C_{q, adjusted}=\log_{2}\!\left(E^{C_{q}}\right)}
#' @param data A data.table containing qPCR data
#' @return A data.table containing the efficiency adjusted Cq values
#'
efficiency_adjustment <- function(data) {
  data[, Cq_adjusted := ifelse(Efficiency < 2, log2(Efficiency ^ Cq), Cq)]
}

# Calculate efficiency adjusted Cq values
efficiency_adjustment(plate_1$samples)
efficiency_adjustment(plate_2$samples)
efficiency_adjustment(plate_3$samples)
efficiency_adjustment(plate_4$samples)

```

## Standard curves

Calculate the standard curve for each qPCR plate.
The standard curve is calculated by fitting a linear model to
$C_q \sim \log_{10} (concentration).

```{r standard_curve}

# Remove dilutions < 10^-5 from standards
plate_1$standards <- plate_1$standards[Concentration > 1e-5]
plate_2$standards <- plate_2$standards[Concentration > 1e-5]
plate_3$standards <- plate_3$standards[Concentration > 1e-5]
plate_4$standards <- plate_4$standards[Concentration > 1e-5]

#' Calculate the standard curve for a qPCR plate
#' @param data A data.table containing qPCR standard curve data
#' @return A list containing the intercept, slope, R^2, and efficiency
#'
calculate_standard_curve <- function(data, adjust_cq = FALSE, use_tcn = FALSE) {
  data[, {
    formula <- if (adjust_cq) {
      if (use_tcn) Cq_adjusted ~ log10_TCN else Cq_adjusted ~ log10_concentration
    } else {
      if (use_tcn) Cq ~ log10_TCN else Cq ~ log10_concentration
    }
    model <- lm(formula, data = .SD)
    intercept  <- model$coefficients[1]
    slope      <- model$coefficients[2]
    r_squared  <- summary(model)$r.squared
    efficiency <- efficiency(slope)
    list(
      Plate      = Plate[1],
      Target     = Target[1],
      N          = .N,
      Intercept  = intercept,
      Slope      = slope,
      R2         = r_squared,
      Efficiency = efficiency
    )
  }, by = .(Target, Type)]
}

# Calculate the standard curve
plate_1$standard_curve <- calculate_standard_curve(plate_1$standards)
plate_2$standard_curve <- calculate_standard_curve(plate_2$standards)
plate_3$standard_curve <- calculate_standard_curve(plate_3$standards)
plate_4$standard_curve <- calculate_standard_curve(plate_4$standards)

kable(plate_1$standard_curve, digits = 2)
kable(plate_2$standard_curve, digits = 2)
kable(plate_3$standard_curve, digits = 2)
kable(plate_4$standard_curve, digits = 2)

```

### Standard efficiency adjusted $C_q$

```{r standard_efficiency_adjustment}

merge_efficiency <- function(data, standard_curve) {
  data[standard_curve, on = .(Target, Type), Efficiency := i.Efficiency]
}

merge_efficiency(plate_1$standards, plate_1$standard_curve)
merge_efficiency(plate_2$standards, plate_2$standard_curve)
merge_efficiency(plate_3$standards, plate_3$standard_curve)
merge_efficiency(plate_4$standards, plate_4$standard_curve)

# Calculate efficiency adjusted Cq values for standards
efficiency_adjustment(plate_1$standards)
efficiency_adjustment(plate_2$standards)
efficiency_adjustment(plate_3$standards)
efficiency_adjustment(plate_4$standards)

```

### Standard curve plots

```{r standard_curve_plots}

#' Plot standard curve
plot_standard_curve <- function(data, standard_curve) {
  types <- unique(data$Type)
  plots <- list()
  for (type in types) {
    plots[[type]] <- ggscatter(
      data[Type == type], x = "log10_concentration", y = "Cq_adjusted",
      add = "reg.line", conf.int = TRUE
    ) +
      stat_regline_equation(
        aes(label = paste(
          after_stat(eq.label), after_stat(rr.label), sep = "~~~~"
        )),
        label.x = -2.5, label.y = 32
      ) +
      ggtitle(paste0("Plate ", data$Plate[1], " ", type, " ", data$Target[1])) +
      xlab(TeX("$\\log_{10}(Concentration)$")) +
      ylab(TeX("$C_q$"))
  }
  ggarrange(plotlist = plots)
}

plot_standard_curve(plate_1$standards, plate_1$standard_curve)
plot_standard_curve(plate_2$standards, plate_2$standard_curve)
plot_standard_curve(plate_3$standards, plate_3$standard_curve)
plot_standard_curve(plate_4$standards, plate_4$standard_curve)

```

```{r combined_standard_curve}

combined_standard <- rbindlist(list(
  plate_1$standards,
  plate_2$standards,
  plate_3$standards,
  plate_4$standards
))

combined_standard_curve <- calculate_standard_curve(combined_standard)

kable(combined_standard_curve, digits = 2)

plot_standard_curve(combined_standard[Target == "ITS"], combined_standard_curve)
plot_standard_curve(combined_standard[Target == "16S"], combined_standard_curve)

bac_initial_copies <- 1e7
fun_initial_copies <- 1e5

# Add TCN to combined standard from dilution range from initial copies based on concentration of standards
combined_standard[Target == "ITS", TCN := Concentration * fun_initial_copies]
combined_standard[Target == "16S", TCN := Concentration * bac_initial_copies]
combined_standard[, log10_TCN := log10(TCN)]

# Plot TCN vs Cq
ggscatter(
  combined_standard, x = "log10_TCN", y = "Cq_adjusted",
  add = "reg.line", conf.int = TRUE
) +
  stat_regline_equation(
    aes(label = paste(
      after_stat(eq.label), after_stat(rr.label), sep = "~~~~"
    )),
    label.x = 2, label.y = 15
  ) +
  facet_grid(Target ~ Type, scales = "free") +
  ggtitle("Combined Standard Curve") +
  xlab(TeX("$\\log_{10}(TCN)$")) +
  ylab(TeX("$C_q$"))

# Plot TCN vs Cq_adjusted for each combination of Target and Type
# ggplot(combined_standard, aes(x = log10(TCN), y = Cq_adjusted)) +
#   geom_point() +
#   geom_smooth(method = "lm", se = TRUE) +
#   stat_regline_equation(
#     aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~~")),
#     label.x = 0.5, label.y = 15
#   ) +
#   facet_grid(Target ~ Type, scales = "free") +
#   ggtitle("Combined Standard Curves") +
#   xlab(TeX("$\\log_{10}(TCN)$")) +
#   ylab(TeX("$C_q$"))

```

### Standard curve copy number

```{r standard_copy_number}

# Calculate TCN for standards
for (plate in list(plate_1, plate_2, plate_3, plate_4)) {
  plate$standards[Target == "ITS", TCN := Concentration * fun_initial_copies]
  plate$standards[Target == "16S", TCN := Concentration * bac_initial_copies]
  plate$standards[, log10_TCN := log10(TCN)]
}

plate_1$standards

# Calculate standard curves for each plate using TCN
for (plate in list(plate_1, plate_2, plate_3, plate_4)) {
  plate$tcn_standard_curve <- calculate_standard_curve(
    plate$standards, adjust_cq = TRUE, use_tcn = TRUE
  )
}

plate_1$tcn_standard_curve <- calculate_standard_curve(
  plate_1$standards, adjust_cq = TRUE, use_tcn = TRUE
)
plate_2$tcn_standard_curve <- calculate_standard_curve(
  plate_2$standards, adjust_cq = TRUE, use_tcn = TRUE
)
plate_3$tcn_standard_curve <- calculate_standard_curve(
  plate_3$standards, adjust_cq = TRUE, use_tcn = TRUE
)
plate_4$tcn_standard_curve <- calculate_standard_curve(
  plate_4$standards, adjust_cq = TRUE, use_tcn = TRUE
)

plate_1$tcn_standard_curve
plate_2$tcn_standard_curve
plate_3$tcn_standard_curve
plate_4$tcn_standard_curve

```

## Repeats

Repeat plate 3 soil ITS due to poor soil standard curve.

```{r repeats}

# Print samples in which all reps have been flagged as outliers
repeats <- function(data) {
  data[
    Content == "Sample",
    .SD[all(Outlier != "ok")],
    by = .(Sample, Target)
  ][
    , .(Plate, Type, Target, Sample, Outlier = "outliers")
  ][
    , .SD[1], by = .(Sample, Target)
  ][
    order(Sample)
  ]
}

kable(repeats(plate_1$data))
kable(repeats(plate_2$data))
kable(repeats(plate_3$data))
kable(repeats(plate_4$data))

```

## Sample copy number

Calculate the theoretical copy number (TCN) of each sample based on the standard curve using the equation:

$$
dilution \times 10^{(C_{q} - intercept)/slope}
$$

```{r sample_copy_number}

#` Calculate TCN for samples
calculate_tcn <- function(data, standard_curve) {
  data[
    standard_curve,
    on = .(Target, Type),
    TCN := Dilution * 10 ^ ((Cq_adjusted - i.Intercept) / i.Slope)
  ]
}

for (plate in list(plate_1, plate_2, plate_3, plate_4)) {
  calculate_tcn(plate$samples, plate$tcn_standard_curve)
}

```

## Sample summary

Calculate the mean and standard deviation of the TCN for each sample.

```{r sample_summary}

#' Calculate sample summary statistics
sample_summary <- function(data) {
  data[
    Content == "Sample",
    .(Mean = mean(TCN), SD = sd(TCN), N = .N),
    by = .(Sample, Target)
  ][
    order(Sample)
  ]
}

sample_summary(plate_1$samples)

ggplot(plate_2$samples, aes(x = Sample, y = log10(TCN))) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  ggtitle("Sample TCN") +
  xlab("Sample") +
  ylab("TCN") +
  theme("ggpubr")

```

## Combined data

```{r combined_data}

combined_data <- rbindlist(list(
  plate_1$samples,
  plate_2$samples,
  plate_3$samples,
  plate_4$samples
))

```

## Repeats

```{r z_score}

# Calculate z score for each replicate
combined_data[, z_score := (TCN - mean(TCN)) / sd(TCN), by = .(Sample, Target)]

combined_data[, z_score := scale(TCN), by = .(Sample, Target)]

combined_data[abs(z_score) > 1.5][order(Target, Sample)]

ggplot(combined_data, aes(x = Sample, y = log10(TCN))) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~ Target, ncol = 1) +
  ggtitle("Sample TCN") +
  xlab("Sample") +
  ylab("TCN") +
  theme("ggpubr")

```

## Summary

Calculate the mean and standard deviation of the TCN for each sample.

Calculate TCN per mg of sample. 2 % of the sample mass was used for DNA extraction.

```{r summary}

summary <- combined_data[,
  .(mean_TCN = mean(TCN), sd_TCN = sd(TCN), N = .N),
  by = .(Target, Sample)
]

# Combined with sample mass
sample_mass <- fread("Data/sample_mass.csv")[, Sample := sprintf("%02d", Sample)]
summary <- merge(summary, sample_mass, by = "Sample")

summary[, TCN_mg := mean_TCN / (mg * 2 / 100)]

kable(summary)

# Export as csv
fwrite(summary, "Data/abundance.csv")

```
